Separating Use and Reuse to Improve Both

### 2017-06-06 FACS/section3.tex
paper

To illustrates how 
\name \emph{improves the use} of classes we model a simplified version of
Set and Bag collections first in Java, and then in \name.
The main benefit of the \name solution is that we can get reuse 
without introducing subtyping between Bags and Sets. This improves the 
use of Bags and eliminates logical errors arizing from incorrect
subtyping relations that are allowed in the Java solution.

@@

\subsection{Sets and Bags in \name}
Instead, in \name, if we was to originally declare
\saveSpace\begin{lstlisting}
Set:{/*set implementation*/} 
\end{lstlisting}\saveSpace
Then our code would be impossible to reuse in the first place for any user of our library.
We consider this an advantage, since unintended code reuse runs into underdocumented behaviour nearly all the time\footnote{See
``Design and document for inheritance or else prohibit
it''~\cite{Bloch08}: the
self use of public methods is rarelly documented, thus is hard to understand the effects of overriding a library metod.
}!
If the designer of the set class wishes to make it reusable, it can do it explicitly by providing a set trait:
\saveSpace\begin{lstlisting}
set:{/*set implementation*/} 
Set:Use set
\end{lstlisting}\saveSpace
Since \Q@set@ can never be used as a type, there is no reason to give it a {\bf fancy-future-aware} name like
\Q@AbstractSetOrBag@.
When bag will be added, the code will look either like
\saveSpace\begin{lstlisting}
set:{/*set implementation*/} 
Set:Use set
Bag: Use set, {/*bag implementation*/}
\end{lstlisting}\saveSpace\saveSpace
or 
\saveSpace\saveSpace\begin{lstlisting}
set:{/*set implementation*/} 
Set:Use set
bag: Use set, {/*bag implementation*/}
Bag: Use bag
\end{lstlisting}\saveSpace
Notice how, thanks to flattening, the resulting code for \Q@Bag@ is identical in both versions, 
and as shown in Section 2, there is no trace of trait \Q@bag@. Thus if we are the developers of bags, we can temporarly go for the first version, and move when needed to the second without adding new undesired complexity for our old clients.

>>>>

To illustrates how 
\name \emph{improves the use} of classes we model a simplified version of
Set and Bag collections first in Java, and then in \name.
The main benefit of the \name solution is that \emph{we can get reuse 
without introducing subtyping between Bags and Sets}. This improves the 
use of Bags and eliminates logical errors arizing from incorrect
subtyping relations that are allowed in the Java solution.

@@

\subsection{Sets and Bags in \name}
Instead, in \name, if we were to originally declare
\saveSpace\begin{lstlisting}
Set:{/*set implementation*/} 
\end{lstlisting}\saveSpace
Then our code would be impossible to reuse in the first place for any user of our library.
We consider this an advantage, since unintended code reuse runs into underdocumented behaviour nearly all the time\footnote{See
``Design and document for inheritance or else prohibit
it''~\cite{Bloch08}: the
self use of public methods is rarelly documented, thus is hard to understand the effects of overriding a library metod.
}!
If the designer of the set class wishes to make it reusable, he/she can do it explicitly by providing a set trait:
\saveSpace\begin{lstlisting}
set:{/*set implementation*/} 
Set:Use set
\end{lstlisting}\saveSpace
Since \Q@set@ can never be used as a type, there is no reason to give it a {\bf fancy-future-aware} name like
\Q@AbstractSetOrBag@.
When bag will be added, the code will look either like
\saveSpace\begin{lstlisting}
set:{/*set implementation*/} 
Set:Use set
Bag: Use set, {/*bag implementation*/}
\end{lstlisting}\saveSpace\saveSpace
or 
\saveSpace\saveSpace\begin{lstlisting}
set:{/*set implementation*/} 
Set:Use set
bag: Use set, {/*bag implementation*/}
Bag: Use bag
\end{lstlisting}\saveSpace
Notice how, thanks to flattening, the resulting code for \Q@Bag@ is identical in both versions, 
and as shown in Section 2, there is no trace of trait \Q@bag@. Thus if we are the developers of bags, we can temporarly go for the first version, and move when needed to the second without adding new undesired complexity for our old clients.

### 2017-06-06 FACS/section4.tex
Revision
paper

\saveSpace\saveSpace\section{Improving Reuse}\saveSpace
\name allows reuse even when subtyping is impossible.
In \name traits do not induce a new (externally visible) type.
However, locally in the trait the programmer can use the special self-type \Q@This@ in order to denote the 
type of \Q@this@.
That is: the program is agnostic to what the \Q@This@ type is going to be, so it can
be later assigned to any (or many) classes. 
The idea is that during flattening, \Q@This@ will be replaced with the actual class name.
In this way, \name allows reuse even when subtyping is
impossible, that is, when a method parameter has type \Q@This@ (also
called a \emph{binary
  method}~\cite{bruce96binary}). 
This type of situations is the primary motivator
for previous work aiming at separating inheritance from subtyping.
Leveraging on the \Q@This@ type, we can also provide self-instantiation (trait methods can create instances of the class using them) and smoothly integrate state and traits: a challenging problem that has limited the flexibility of traits and
reuse in the past.
%First we show how self-instantiation and state are already avaible in the simplest trait model,
%then we show how to radically improve the

@@

\paragraph{Modeling Points} Consider, for example, two simple 
traits that deal with \emph{point} objects. That is, points
in the cartesian plane (with coordinates \lstinline{x} and
\lstinline{y}). The first trait provides a \emph{binary method} that 
sums the point object with another point to return a new point. 
The second trait provides a similar operation that does multiplication 
instead.
\saveSpace 
\begin{lstlisting}
  pointSum: { method int x()  method int y()//getters
    class method This of(int x,int y)//factory method
    method This sum(This that)
      This.of(this.x()+that.x(),this.y()+that.y())//self instantiation
    }
  pointMul: { method int x() method int y()//repeating getters
    class method This of(int x,int y)//repeating factory
    method This mul(This that)
      This.of(this.x()*that.x(),this.y()*that.y())
    }
\end{lstlisting}
\saveSpace
\noindent As we can see, all the state operations (the getters for the 
\lstinline{x} and \lstinline{y} coordinates) are represented as {\bf abstract} methods.
Notice the abstract \Q@class method This of(..)@ which acts as a constructor
for points:
a class method is similar to a \Q@static@ method in Java but can be abstract. 
As for instance methods, they are late bound:  flattening can provide an implementation for them.
Abstract class methods are very similar to the original concept of member functions in the module composition setting.

@@

\noindent However \emph{in a traditional trait model} this would fail since methods \Q@x@,\Q@y@ and \Q@of@ are still abstract!
Instead a user could write something similar to:

@@

\item {\bf Disadvantages:} Writing those obvious definitions to close
  the state/fixpoint in the class 
   with the constructors and fields and getter/setters and factories can be quite tedious.
   %Moreover, there is no way for a trait to specify a default value for a field,
   %the class need to handle all the state, even state that is conceptually
   %"private" of such trait. 
   Previous work shows that such code can be automatically generated~\cite{wang2016classless}.

@@

\paragraph{Points in \name:}
In \name and with our approach to handle the state, 
\lstinline{pointSum} and \lstinline{pointMul} can indeed be directly composed:
\saveSpace
\begin{lstlisting}
  PointAlgebra:Use pointSum,pointMul //coherent class completion!
\end{lstlisting}  
\saveSpace
\noindent
  Note how we can declare the methods independently and compose the result
  as we wish.

@@

In the following you can see how to mix colors and flavours. The syntax \Q@_2merge@ and   \Q@_3merge@
call the version of merge as defined in the second/third element of \use.
\saveSpace\begin{lstlisting}
  p: {/*as before*/ }
  pointSum:Use p, {/*as before*/ }
  colored:{/*as before*/}
  flavored:{method Flavor flavor() //very similar to colored
    method This withFlavor(Flavor that)
    method This merge(This that) //how to merge flavors handled here
      this.withFlavor(that.flavor())}//here we just inherit ``that'' flavor
  FCPoint:Use pointSum,colored,flavored{
    class method This of(int x,int y)
      This.of(x,y,Color.of(/*red*/),Flavor.none())
    class method This of(int x, int y,Color color,Flavor flavor)
    //we can resolve the conflict about two implementations for merge
    //by proving our own implementation here
    method This merge(This that) this._2merge(that)._3merge(that)
    }
\end{lstlisting}  \saveSpace\saveSpace

>>>>

\saveSpace\saveSpace\section{Improving Reuse}\saveSpace
\name allows reuse even when subtyping is impossible.
\name traits do not induce a new (externally visible) type.
However, locally in a trait, programmers can use the special self-type \Q@This@~\cite{bruce_1994,Saito:2009,ryu16ThisType} in order to denote the 
type of \Q@this@.
That is, a program is agnostic to what the \Q@This@ type is, so that it can
be later assigned to any (or many) classes. 
The idea is that during flattening, \Q@This@ will be replaced with the actual class name.
In this way, \name allows reuse even when subtyping is
impossible. For example for \emph{binary
  methods}~\cite{bruce96binary} where the method parameter has type \Q@This@. 
This type of situations is the primary motivator
for previous work aiming at separating inheritance from subtyping~\cite{cook}.
Leveraging on the \Q@This@ type, we can also provide self-instantiation (trait methods can create instances of the class using them) and smoothly integrate state and traits: a challenging problem that has limited the flexibility of traits and
reuse in the past.
%First we show how self-instantiation and state are already avaible in the simplest trait model,
%then we show how to radically improve the

@@

\paragraph{Modeling Points} Consider, for example, two simple 
traits that deal with \emph{point} objects. That is, points
in the cartesian plane (with coordinates \lstinline{x} and
\lstinline{y}). The first trait provides a \emph{binary method} that 
sums the point object with another point to return a new point. 
The second trait provides a similar operation that does multiplication 
instead.
\saveSpace 
\begin{lstlisting}
  pointSum: { method int x()  method int y()//getters
    class method This of(int x,int y)//factory method
    method This sum(This that)
      This.of(this.x()+that.x(),this.y()+that.y())//self instantiation
    }
  pointMul: { method int x() method int y()//repeating getters
    class method This of(int x,int y)//repeating factory
    method This mul(This that)
      This.of(this.x()*that.x(),this.y()*that.y())
    }
\end{lstlisting}
\saveSpace
\noindent As we can see, all the state operations (the getters for the 
\lstinline{x} and \lstinline{y} coordinates) are represented as {\bf abstract} methods.
Notice the abstract \Q@class method This of(..)@ which acts as a constructor
for points:
a class method is similar to a \Q@static@ method in Java but can be abstract. 
As for instance methods, they are late bound:  flattening can provide an implementation for them.
Abstract class methods are very similar to the original concept of member functions in the module composition setting~\cite{ancona2002calculus}.

@@

\noindent However \emph{in a traditional trait model}~\cite{Traits:ECOOP2003} this would fail since methods \Q@x@,\Q@y@ and \Q@of@ are still abstract!
Instead a user could write something similar to:

@@

\item {\bf Disadvantages:} Writing such obvious definitions to close
  the state/fixpoint in the class 
   with the constructors and fields and getter/setters and factories is tedious.
   %Moreover, there is no way for a trait to specify a default value for a field,
   %the class need to handle all the state, even state that is conceptually
   %"private" of such trait. 
   Previous work shows that such code can be automatically
   generated~\cite{wang2016classless}.
   Also a model with fields and constructors is necessarily
   more complex than a model with methods only.

@@

\paragraph{Points in \name:}
In \name and with our approach to handle the state, 
\lstinline{pointSum} and \lstinline{pointMul} can indeed be directly composed:
\saveSpace
\begin{lstlisting}
  //Same code as before! Works because resulting class is coherent.
  PointAlgebra:Use pointSum,pointMul 
\end{lstlisting}  
\saveSpace
\noindent
  Note how we can declare the methods independently and compose the result
  as we wish.

@@

The following code shows how to mix colors and flavours. The syntax \Q@_2merge@ and   \Q@_3merge@
call the version of merge as defined in the second/third element of \use.
\saveSpace\begin{lstlisting}
  p: {/*as before*/ }
  pointSum:Use p, {/*as before*/ }
  colored:{/*as before*/}
  flavored:{method Flavor flavor() //very similar to colored
    method This withFlavor(Flavor that)
    method This merge(This that) //merging flavors handled here
      this.withFlavor(that.flavor())}//inherits ``that'' flavor
  FCPoint:Use pointSum,colored,flavored{
    class method This of(int x,int y)
      This.of(x,y,Color.of(/*red*/),Flavor.none())
    class method This of(int x, int y,Color color,Flavor flavor)
    //resolves the conflict about two implementations for merge
    //by proving our own implementation here
    method This merge(This that) this._2merge(that)._3merge(that)
    }
\end{lstlisting}  \saveSpace\saveSpace
